---
title: "Cheatsheet TDDE01"
author: "Rolf Sievert (rolsi701)"
header-includes:
    - \usepackage[final]{pdfpages}
---

# Miscellaneous

## Vocabulary
Generalization error: error on test data.

## Info
Test data is after validation, validation data is during testing of models.

Naïve Bayes: doesn't combine information to classify, tries to classify with individual properties. Doesn't work well with linear separations.

Gets biggest value:

```{r, eval=FALSE}
which_max(data)
```

How to read R scripts:

*{r, code = readLines("lab1/ass2.R"), eval=FALSE}*

Read data and split into random sets

```{r eval=FALSE}
video = read.csv("video.csv")
data = subset(video, select=-utime)
data_size = dim(data)[1]
set.seed(12345)
i1 = sample(1:data_size, floor(data_size*0.5))
i2 = setdiff(1:data_size, i1)
train = data[i1,]
test = data[i2,]
```

## Plotting
```{r eval=FALSE}
library("ggplot2")
library("gridExtra")
plot.projection = ggplot() +
geom_point(data=spectra, mapping=aes(x=PC1, y=PC2, color=class))
grid.arrange(plot.projection)
```

```{r eval=FALSE}
plot(data, type='l', col="red", ylim=c(100, 500), ylab="Error")
plot(data2, type='l', col="blue")
```

## PCA and select principal components

```{r eval=FALSE}
library("stats")
pca.fit = prcomp(data, scale.=T)
# screeplot(pca.fit) to get all eigen values
# Get eigen values (standard deviation ^ 2)
lambda = pca.fit\$sdev^2
sprintf("%2.3f", lambda/sum(lambda))
PC1 = pca.fit\$x[,1]
PC2 = pca.fit\$x[,2]
```

Use *princomp* for loadings to get eigenvectors.

```{r eval=FALSE}
library("stats")
pca.fit = princomp(data)
loadings(pca.fit)
```

## Confusion matrix

```{r eval=FALSE}
t <- table(prediction, true_values)
(t[1,2]+t[2,1])/sum(t)
```

# Labs

## Lab 1

``` {r child="lab1/lab1.Rmd"}
```

## Lab 2

``` {r child="lab2/lab2.Rmd"}
```

## Lab 3

``` {r child="lab3/Rolf_Sievert.Rmd"}
```

# Other

```{r, code = readLines("exams/2017-04-18/tenta_crabs.R"), eval=FALSE}
```

```{r, code = readLines("exams/2017-04-18/tenta_bank.R"), eval=FALSE}
```

```{r, code = readLines("exams/2018-01-11/tenta_1.R"), eval=FALSE}
```

```{r, code = readLines("exams/2018-01-11/tenta_2.R"), eval=FALSE}
```

```{r, code = readLines("exams/2018-01-11/tenta_2_2.R"), eval=FALSE}
```

\includepdf[pages=-]{lectureslides_shorter.pdf}

\includepdf[pages=-, landscape=true]{rmarkdown-cheatsheet.pdf}

\includepdf[pages=-, landscape=true]{extra.pdf}
